An Accurate Model for SARS-CoV-2 Pooled RT-PCR Test Errors

### 2021-09-09 paper.tex
with some review corrections

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development of techniques to increase the the
throughput of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30,
  BayesianDorfman, MatrixPooling, LionDorfman}. These techniques,
often termed \emph{pooling}, or group testing, have originated in the
seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroguhly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

@@

Below, we refer to \eqref{eq:common} as the \emph{common
  assumption}.

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume a more
realistic model: amplification of viral RNA succeeds or fails for each
sample independently. Furthermore, according to \cite{Simplistic1,
  Simplistic2, Kim, OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool either. For lack of
data pointing otherwise, we incorporate this assumption into our model
with a small modification. We do assume that a false amplification can
occur only once per pool. However, we also assume false amplification
is independent of any other correct amplification. Specifically, it is
possible that every correct amplification fails \emph{and} an
erroneous one occurs simultaneously. This assumption is somewhat
specific for the current application of screening for SARS-CoV-2 via
RT-PCR. For example, cross-reactivity with other coronaviruses would
have violated this assumption. However, cross-reactivity was ruled out
in \cite{KitComparison}. These assumptions lead to our model, which is
illustrated in Figure~\ref{fig:likelihood} and summarized in
\eqref{eq:likelihood}.
\begin{align}\label{eq:likelihood}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N (1-\sens)^{\theta_i} =
    \spec(1-\Se)^{\sum_i\theta_i}.
\end{align}

@@

\begin{align}
  \begin{split}
    \Pr(\data=0|\theta_1=1) &= \sum_{\theta_2,\dots,\theta_N}
    \Pr(\theta_2,\dots,\theta_N)
    \Pr(\data=0|\theta_1=1,\theta_2,\dots,\theta_N) \\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\Pr(\sum_{i=2}^N\theta_i=k)
    \Pr(\data=0|\theta_1=1,\sum_{i=2}^N\theta_i = k)\\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\binom{N-1}{k}q^k(1-q)^{N-1-k} \Sp(1-\Se)^{1+k}\\
    %
    %
    %
    &= \Sp(1-\Se) \sum_{k=0}^{N-1}\binom{N-1}{k}
    \left(q(1-\Se)\right)^k(1-q)^{N-1-k}\\
    %
    %
    %
    &= \Sp(1-\Se)(1-q\Se)^{N-1}. \\
  \end{split}
\end{align}

@@

\subsubsection*{Comparison metric}
The single test false-negative rate $1-\Se$ and scheme false-negative
rate $\mi$ are compared via:
\begin{equation}\label{eq:erel}
\err{exact} := \frac{\mi - (1-\Se)}{1-\Se} \cdot 100.
\end{equation}
$\err{exact}$ is the percentage increase in the pooling scheme
false-negative rate, relative to the single test false-negative rate.

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A straight forward calculation shows that this implies the
percentage increase in scheme false-negative rate is $\err{common} :=
100\cdot \Se$.

@@

\section*{Results}\label{section:results}
\subsection*{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\err{exact}$ can be expected (Figure \ref{fig1}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure~\ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure~\ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative to
    the single test false-negative rates \eqref{eq:erel}. Right:
    colors represent the difference between $\err{common}$ and
    $\err{exact}$. The disease prevalence $q$, is varied on the
    x-axis, while the test sensitivity is varied on the y-axis. Pool
    size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}. The left panel shows that $\err{exact}$ is
    largest for low prevalence values $q$. The difference between
    $\err{common}$ and $\err{exact}$ can be as large as 30\%, as seen
    in the right panel. }\label{fig1}
\end{figure}

@@

Thus, improving pooling schemes is imperative. An especially important
consideration in designing such schemes is explicitly taking the
intrinsic PCR error rates into account. We \cite{DOPE}, as well as
others \cite{BayesianDorfman}, have developed such models, although
they have not yet been implemented in real-world settings.

>>>>

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development of techniques to increase the throughput
of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman,
  MatrixPooling, LionDorfman}. These techniques, often termed
\emph{pooling}, or group testing, have originated in the seminal work
of Dorfman and his pooling technique \cite{DorfmanOriginal,
  DorfmanYuvalDor}. Under Dorfman pooling, one selects $N$ individuals
and performs a single RT-PCR test on their combined (\emph{pooled})
samples. If the pooled test yields a positive result, then each
individual is retested separately; otherwise, everyone is declared
negative. The throughput efficiency of Dorfman pooling has been
demonstrated empirically \cite{DorfmanYuvalDor} and its error rates
thoroughly investigated \cite{Kim, Simplistic1, OptimalDorfmanPool}.

@@

Below, we refer to equation \eqref{eq:common} as the \emph{common
  assumption}.

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume a more
realistic model: amplification of viral RNA succeeds or fails for each
sample independently. Furthermore, according to \cite{Simplistic1,
  Simplistic2, Kim, OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool either. For lack of
data pointing otherwise, we incorporate this assumption into our model
with a small modification. We do assume that a false amplification can
occur only once per pool. However, we also assume false amplification
is independent of any other correct amplification. Specifically, it is
possible that every correct amplification fails \emph{and} an
erroneous one occurs simultaneously. This assumption is somewhat
specific for the current application of screening for SARS-CoV-2 via
RT-PCR. For example, cross-reactivity with other coronaviruses would
have violated this assumption. However, cross-reactivity was ruled out
in \cite{KitComparison}. These assumptions lead to our model, which is
illustrated in Figure~\ref{fig:likelihood} and summarized in
equation \eqref{eq:likelihood}.
\begin{align}\label{eq:likelihood}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N (1-\sens)^{\theta_i} =
    \spec(1-\Se)^{\sum_i\theta_i}.
\end{align}

@@

\begin{align}
  \begin{split}
    \Pr(\data=0|\theta_1=1) &= \sum_{\theta_2,\dots,\theta_N}
    \Pr(\theta_2,\dots,\theta_N)
    \Pr(\data=0|\theta_1=1,\theta_2,\dots,\theta_N) \\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\Pr\left(\sum_{i=2}^N\theta_i=k\right)
    \Pr\left(\data=0|\theta_1=1,\sum_{i=2}^N\theta_i = k\right)\\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\binom{N-1}{k}q^k(1-q)^{N-1-k} \Sp(1-\Se)^{1+k}\\
    %
    %
    %
    &= \Sp(1-\Se) \sum_{k=0}^{N-1}\binom{N-1}{k}
    \left(q(1-\Se)\right)^k(1-q)^{N-1-k}\\
    %
    %
    %
    &= \Sp(1-\Se)(1-q\Se)^{N-1}. \\
  \end{split}
\end{align}

@@

\subsubsection*{Comparison metric}
The single test false-negative rate $1-\Se$ and scheme false-negative
rate $\mi$ are compared via:
\begin{equation}\label{eq:erel}
\err{exact} := \frac{\mi - (1-\Se)}{1-\Se} \cdot 100\%.
\end{equation}
$\err{exact}$ is the percentage increase in the pooling scheme
false-negative rate, relative to the single test false-negative rate.

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A straight forward calculation shows that this implies the
percentage increase in scheme false-negative rate is $\err{common} :=
100\%\cdot \Se$.

@@

\section*{Results}\label{section:results}
\subsection*{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, the increase in scheme false-negative rate
(compared to $1-\Se$), summarized in $\err{exact}$, is at leaset
$60\%$ (Figure \ref{fig1}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from equation \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure~\ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure~\ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative to
    the single test false-negative rates equation \eqref{eq:erel}. Right:
    colors represent the difference between $\err{common}$ and
    $\err{exact}$. The disease prevalence $q$, is varied on the
    x-axis, while the test sensitivity is varied on the y-axis. Pool
    size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}. The left panel shows that $\err{exact}$ is
    largest for low prevalence values $q$. The difference between
    $\err{common}$ and $\err{exact}$ can be as large as 30\%, as seen
    in the right panel. }\label{fig1}
\end{figure}

@@

Thus, improving pooling schemes is imperative. An especially important
consideration in designing such schemes is explicitly taking the
intrinsic PCR error rates into account. As we have demonstrated above,
analyzing expected error rates of a given scheme, given some error
model is relatively straightforward. Designing a pooling scheme
tailored for a specific error model is, however, far from trivial. In
a recent study \cite{DOPE}, we have developed a novel pooling scheme
that takes advantage of this specific error model. Our new pooling
scheme shows great promise in terms of both accuracy, as well as
efficiency, although it has not yet been implemented in real-world
settings.
